An open reproducible framework for the study of the iterated prisoner's dilemma

### 2016-06-02 tex/paper.tex
Most smaller comments now addressed.
Still need work on library structure and/or reuse section.

Several Iterated Prisoner's Dilemma tournaments have generated much interest;
including Axelrod's original tournaments \cite{Axelrod1980a,Axelrod1980b},
two 2004 anniversary tournaments \cite{kendall2007iterated}, and
the Stewart and Plotkin 2012 tournament \cite{Stewart2012}, following the
discovery of zero-determinant strategies.  Subsequent research has spawned an
enormous number of papers, but rarely are the results reproducible. Among
well-known tournaments, in only one case is the full original source code
available (Axelrod's second tournament \cite{Axelrod1980b}, in FORTRAN). In no
cases is the available code well-documented, easily modifiable, or released with
significant test suites.

@@

This paper introduces a software package: the Axelrod Python library
\cite{Axelrod-Pythonprojectteam2015}. The Axelrod-Python project has the
following stated goals:

@@

\begin{itemize}[noitemsep,topsep=0pt]
    \item Open: all code is released under an MIT license;
    \item Reproducible and well-tested: at the time of writing there is an excellent level of
        integrated tests with 99.37\% coverage (including property based tests:
        \cite{Hypothesis3.0.3}
    \item Well-documented: all features of the library are documented for ease of
        use and modification
    \item Extensive: 123 strategies are included, with infinitely-many
        strategies available in the case of parametrised strategies
    \item Extensible: easy to modify to include new strategies and to run new tournaments
\end{itemize}

@@

The game of~(\ref{equ:one-shot}) is called the Prisoner's Dilemma. Specific
numerical values of \((R,S,T,P)=(3,0,5,1)\) are often used in the literature,
although any satisfying the conditions in~\ref{equ:one-shot} will yield similar
results.  Axelrod's tournaments (and further implementations of these) are
sometimes referred to as Iterated Prisoner's Dilemma (IPD) tournaments. An
incomplete overview of published tournaments is given in
Table~\ref{tab:tournaments}.

\begin{table}[!hbtp]
    \begin{center}
        \begin{tabular}{ccccc}
            \toprule
            Year     & Reference                  & Number of Strategies & Type     & Source Code\\
            \midrule
            1979     & \cite{Axelrod1980a}        & 13                   & Standard & Not immediately available\\
            1979     & \cite{Axelrod1980b}        & 64                   & Standard & Available in FORTRAN\\
            1991     & \cite{Bendor1991}          & 13                   & Noisy    & Not immediately available\\
            2002     & \cite{Stephens2002}        & 16                   & Wildlife & Not applicable\\
            2005     & \cite{kendall2007iterated} & 223                  & Varied   & Not available \\
            2012     & \cite{Stewart2012}         & 13                   & Standard & Not fully available \\
            \bottomrule
        \end{tabular}
    \end{center}
    \caption{An overview of a selection of published tournaments. Not all
             tournaments were `standard' round robins; for more details
             see the indicated references.}\label{tab:tournaments}
\end{table}

@@

Installation of the library is straightforward as it is available via the
standard Python installation package: `pip'
(\url{https://pypi.python.org/pypi}). This ensures it can be used on all major
operating systems (Windows, OS X and Linux).

@@

Figure~\ref{fig:demo_tournament_commands} shows the very basic utilisation
of the library and further details can be found at the online documentation:
\url{http://axelrod.readthedocs.org}.

@@

To date the library has had contributions from 24 contributors from a variety
of backgrounds. These contributions have been in terms of strategies. One
strategy is the creation of an undergraduate mathematics student with little
prior knowledge of programming. Multiple other strategies were written by a 15
year old secondary school student. Both of these students are authors of this
paper. As well as these strategy contributions, vital architectural
improvements to the library itself have also been received. You can see an
overview of the structure of the source code in Figure~\ref{fig:overview}.

@@

This can be reproduced as shown in Figure~\ref{fig:stewart-code}, which gives
the plot of Figure~\ref{fig:stewart_tournament}.

@@

The current winning strategy is a novel strategy called: Looker Up. This is a
strategy that maps a given set of states to actions. The state space is defined
generically by \(m, n\) so as to map states to actions as shown
in~(\ref{equ:example_of_lookerup}).

@@

The example of (\ref{equ:example_of_lookerup}) is an incomplete illustration of
the mapping for \(m=8, n=2\). Intuitively, this state space uses the initial
plays of the opponent to gain some information about its intentions whilst still
taking into account the recent play. The actual winning strategy is an instance
of the framework for \(m=n=2\) for which a particle swarm algorithm has been
used to train it, the second placed strategy was trained with an evolutionary
algorithms. More details of this can be found in \cite{Jones2015, George2016}.
In \cite{Knight2015} experiments are described that evaluate how the second
placed strategy behaves in environments other than those in which it was trained
and it continues to perform strongly.

There are various other insights that have been gained from ongoing open
research on the library. These include:

\begin{itemize}[noitemsep,topsep=0pt]
    \item A closer look at zero determinant strategies, showing that
        extortionate strategies obtain a large number of wins \textit{but} do
        not perform particularly well. This is relevant given the findings of
        \cite{Stewart2012} in which zero determinant strategies are shown to be
        able to perform better than any other strategy. This finding extends
        to noisy tournaments (which are also implemented in the library).
    \item This negative relationship between wins and performance does not
        generalise. There are some strategies that perform well, both in terms
        of matches won and overall performance: Back stabber, Double crosser,
        Looker Up, and Fool Me Once. These strategies continue to perform well in noisy
        tournaments, however some of these have knowledge of the length of the
        game (Back stabber and Double crosser). This is not necessary to rank
        well in both wins and score as demonstrated by Looker Up and Fool Me
        Once.
    \item Strategies like Looker Up and Meta Hunter seem to be generally
        cooperative yet still exploit naive strategies. The MetaHunter strategy
        is a particular type of Meta strategy which uses a variety of other
        strategy behaviours to choose a best action. These strategies perform
        very well in general and continue to do so in noisy tournaments.
\end{itemize}

Details for the above can be found in \cite{Harper2015}.

@@

The Axelrod library runs on all major operating systems: Linux, Mac OSX and
Windows.

@@

The library is continuously tested for Python 2.7, 3.3, 3.4 and 3.5.

@@

\begin{multicols}{3}
    \begin{itemize}[noitemsep,topsep=0pt]
        \item Numpy 1.9.2:
        \item Matplotlib 1.4.2:
        \item Cloudpickle  0.1.1:
        \item Hypothesis  3.0:
        \item Testfixtures  4.9.1:
    \end{itemize}
\end{multicols}

@@

The names of all the contributors are not known: as these were mainly done
through Github and certain have not provided their name or responded to a
request for further details. Here is an incomplete list:

@@

The ability to have readily available a large number of strategies makes this
tool an excellent and obvious example of the benefits of open research which
should positively impact the game theoretic community.
This has been evidenced by the work described: the library has already shown
it's potential for reuse.

>>>>

Several Iterated Prisoner's Dilemma tournaments have generated much interest;
Axelrod's original tournaments \cite{Axelrod1980a,Axelrod1980b}, two 2004
anniversary tournaments \cite{kendall2007iterated}, and the Stewart and Plotkin
2012 tournament \cite{Stewart2012}, following the discovery of zero-determinant
strategies.  Subsequent research has spawned a number of papers (many of which
are referenced throughout this paper), but rarely are the results reproducible.
Amongst well-known tournaments, in only one case is the full original source code
available (Axelrod's second tournament \cite{Axelrod1980b}, in FORTRAN). In no
cases is the available code well-documented, easily modifiable, or released
with significant test suites.

@@

This paper introduces a software package: the Axelrod-Python library
\cite{Axelrod-Pythonprojectteam2015}. The Axelrod-Python project has the
following stated goals:

@@

\begin{itemize}[noitemsep,topsep=0pt]
    \item Open: all code is released under an MIT license;
    \item Reproducible and well-tested: at the time of writing there is an excellent level of
        integrated tests with 99.37\% coverage (including property based tests:
        \cite{Hypothesis3.0.3})
    \item Well-documented: all features of the library are documented for ease of
        use and modification
    \item Extensive: 123 strategies are included, with infinitely-many
        strategies available in the case of parametrised strategies
    \item Extensible: easy to modify to include new strategies and to run new tournaments
\end{itemize}

@@

The game of~(\ref{equ:one-shot}) is called the Prisoner's Dilemma. Specific
numerical values of \((R,S,T,P)=(3,0,5,1)\) are often used in the literature
\cite{Axelrod1980a, Axelrod1980b}, although any satisfying the conditions
in~\ref{equ:one-shot} will yield similar results. Axelrod's tournaments (and
further implementations of these) are sometimes referred to as Iterated
Prisoner's Dilemma (IPD) tournaments. An incomplete representative overview of
published tournaments is given in Table~\ref{tab:tournaments}.

\begin{table}[!hbtp]
    \begin{center}
        \begin{tabular}{ccccc}
            \toprule
            Year     & Reference                  & Number of Strategies & Type     & Source Code\\
            \midrule
            1979     & \cite{Axelrod1980a}        & 13                   & Standard & Not immediately available\\
            1979     & \cite{Axelrod1980b}        & 64                   & Standard & Available in FORTRAN\\
            1991     & \cite{Bendor1991}          & 13                   & Noisy    & Not immediately available\\
            2002     & \cite{Stephens2002}        & 16                   & Wildlife & Not a computer based tournament\\
            2005     & \cite{kendall2007iterated} & 223                  & Varied   & Not available \\
            2012     & \cite{Stewart2012}         & 13                   & Standard & Not fully available \\
            \bottomrule
        \end{tabular}
    \end{center}
    \caption{An overview of a selection of published tournaments. Not all
             tournaments were `standard' round robins; for more details
             see the indicated references.}\label{tab:tournaments}
\end{table}

@@

Installation of the library is straightforward as it is available via the
standard Python installation package: `pip'
(\url{https://pypi.python.org/pypi}). The package name is \texttt{axelrod} and
can thus be installed by calling: \texttt{pip install axelrod}. This ensures it
can be used on all major operating systems (Windows, OS X and Linux).

@@

@@

To date the library has had contributions from 24 contributors from a variety
of backgrounds which are not solely academic. These contributions have been in
terms of strategies. One strategy is the creation of an undergraduate
mathematics student with little prior knowledge of programming. Multiple other
strategies were written by a 15 year old secondary school student. Both of
these students are authors of this paper. As well as these strategy
contributions, vital architectural improvements to the library itself have also
been received.

You can see an overview of the structure of the source code in
Figure~\ref{fig:overview}. Add:

\begin{itemize}
	\item Discuss match generators (possibly include in image?)
	\item Discuss what a strategy is (class, inheritance).
	\item Reuse (possibility of creating new tournament types).
\end{itemize}

@@

This can be reproduced as shown in Figure~\ref{fig:stewart-code}, which gives
the plot of Figure~\ref{fig:stewart_tournament}. Note that slight differences
with the results of \cite{Stewart2012} are due to stochastic behaviour.

@@

The current winning strategy is a novel strategy: Looker Up. This is a
strategy that maps a given set of states to actions. The state space is defined
generically by \(m, n\) so as to map states to actions as shown
in~(\ref{equ:example_of_lookerup}).

@@

The example of (\ref{equ:example_of_lookerup}) is an incomplete illustration of
the mapping for \(m=8, n=2\). Intuitively, this state space uses the initial
plays of the opponent to gain some information about its intentions whilst still
taking into account the recent play. The actual winning strategy is an instance
of the framework for \(m=n=2\) for which a particle swarm algorithm has been
used to train it. The second placed strategy was trained with an evolutionary
algorithm \cite{Jones2015, George2016}.
In \cite{Knight2015} experiments are described that evaluate how the second
placed strategy behaves in environments other than those in which it was trained
and it continues to perform strongly.

There are various other insights that have been gained from ongoing open
research on the library, details can be found in \cite{Harper2015}. These include:

\begin{itemize}[noitemsep,topsep=0pt]
    \item A closer look at zero determinant strategies, showing that
		extortionate strategies obtain a large number of wins: the number of
		times they outscore an opponent during a given match. \textit{However} these do
		not perform particularly well from the overall tournament ranking point of
		view. This is relevant given the findings of \cite{Stewart2012} in which zero
		determinant strategies are shown to be able to perform better than any other
		strategy. This finding extends to noisy tournaments (which are also implemented
		in the library).
    \item This negative relationship between wins and performance does not
        generalise. There are some strategies that perform well, both in terms
        of matches won and overall performance: Back stabber, Double crosser,
        Looker Up, and Fool Me Once. These strategies continue to perform well in noisy
        tournaments, however some of these have knowledge of the length of the
        game (Back stabber and Double crosser). This is not necessary to rank
        well in both wins and score as demonstrated by Looker Up and Fool Me
        Once.
    \item Strategies like Looker Up and Meta Hunter seem to be generally
        cooperative yet still exploit naive strategies. The Meta Hunter strategy
        is a particular type of Meta strategy which uses a variety of other
        strategy behaviours to choose a best action. These strategies perform
        very well in general and continue to do so in noisy tournaments.
\end{itemize}

@@

The Axelrod library runs on all major operating systems: Linux, Mac OS X and
Windows.

@@

The library is continuously tested for compatibility with Python 2.7, 3.4 and 3.5.

@@

\begin{multicols}{3}
    \begin{itemize}[noitemsep,topsep=0pt]
        \item Numpy 1.9.2:
        \item Matplotlib 1.4.2:
        \item Hypothesis  3.0:
        \item Testfixtures  4.9.1:
    \end{itemize}
\end{multicols}

@@

The names of all the contributors are not known: as these were mainly done
through Github and some have not provided their name or responded to a
request for further details. Here is an incomplete list:

@@

The ability to have readily available a large number of strategies makes this
tool an excellent and obvious example of the benefits of open research which
should positively impact the game theoretic community.
This has been evidenced by the work described: the library has already shown
its potential for reuse.

Add more here: point to previous discussion about new strategies, new
tournaments, tests etc\dots
